Embedding Intuitionistic into Classical Logic

### 2022-07-16 CSL/lipics-v2021-EIICL.tex
Incorporated Remarks, fixed Lemma 41

\begin{definition}
Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^\circ$.
We say $\psi\in\mathcal F_\to$ is \emph{fulfilled} at $u\in M$ iff $A_\psi^I(u)\to B_\psi^I(u)$ is false or $C_\psi^I(u)$ is true. For $\psi\in\mathcal F_\to$ define $$g_\psi : M\to M, u\mapsto\begin{cases}
		u,&\text{ if $\psi$ is fulfilled in $u$,}\\
		f^I_\psi(u),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal F_\to$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $\psi_1\dots\psi_n$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}(\psi_1\dots \psi_n) = P^I(g_{\psi_n}(\dots(g_{\psi_1}(s^I))\dots)).$$
	\end{itemize}
\end{definition}
\begin{lemma}~\label{thm:prop-countermodel-reduction}
	Let $\mathcal M = (M, I)$ be a counter-model to $\mathcal \varphi^\circ$.
	\begin{enumerate}
		\item $\psi$ is fulfilled at $f_\psi^I(u)$ for all $u\in M, \psi\in\mathcal F_\to$.
		\item If $\psi$ is fulfilled at some $u\in M$ then $\psi$ is fulfilled at all $w\geq u$.
	\end{enumerate}
\end{lemma}

@@

Using $E$ we then obtain $\mathcal S^\circ$ analogously to the predicate case by including formulas:

@@

Then $\varphi = \bigwedge\mathcal S\to P$ is intuitionistically valid if and only if
$$\varphi^\circ= \text{World}(s)\to K(\varphi)\to \bigwedge S^\circ\to P^\#(s)$$
is classically valid, where $s$ is a new constant symbol. This was obvious in the propositional case but here it is more nuanced. A proof is given in the appendix as Lemma~\ref{lemma:fo-simplification}.

@@

As in the predicate case we now show the existence of a canonical counter-model (in case such a counter-model exists). Denote the set of formulas $\psi\in\mathcal S$ of the form $\forall\vec z((A(\vec a)\to B(\vec b))\to C(\vec c))$ with $\mathcal F_\to$ and of the form $\forall \vec z(\forall xA(\vec a)\to B(\vec b))$ with $\mathcal F_\forall$. Let $\mathcal F = \mathcal F_\to\cup\mathcal F_\forall$. For each  $\psi\in\mathcal F$ consider a new function symbol $f_\psi$. As before we may assume that $s$ is a least element and we can avoid the quantification over $w$ by considering the formulas
\begin{itemize}
	\item $\forall \vec z\forall u(\vec E(\vec z, u)\to (A^\#(\vec a, f_\psi(\vec z, u))\to B^\#(\vec b, f_\psi(\vec z, u)))\to C^\#(\vec c, u))$\\instead of $\forall \vec z\forall u(s\preceq u\to\vec E(\vec z, u)\to \forall w(u\preceq w\to A^\#(\vec a, w)\to B^\#(\vec b, w))\to C^\#(\vec c, u))$,
	\item $\forall \vec z\forall u(\vec E(\vec z, u)\to (\forall x(E(x, f_\psi(\vec z, u))\to A^\#(\vec a, f_\psi(\vec z, u))))\to B^\#(\vec b, u))$\\instead of $\forall \vec z\forall u(s\preceq u\to\vec E(\vec z, u)\to \forall w(u\preceq w\to \forall x(E(x, w)\to A^\#(\vec a, w)))\to B^\#(\vec b, u))$,
\end{itemize}
and by removing removing the $s\preceq u$ for the remaining formulas in $\mathcal S^\circ$.
We denote the resulting set of formulas by $\mathcal S^\#$.
Then, the previous formulas are equivalid to
$$\varphi^\circ= \forall u(s\preceq u)\to \bigwedge\{\forall \vec t\forall u(u\preceq f(\vec z, u))\:|\:\psi\in\mathcal F\}\to K(\varphi)\to\bigwedge\mathcal S^\#\to P^\#(s).$$
\todo{we already define $\varphi^\circ$, shouldn't this be some $\varphi^\dagger$? Also the phrase before should be: Then,$\varphi^\circ$ is equivalid to ...?}
As in the previous section we can define a tree counter-model for this.

\begin{definition}
	Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^\circ$. \todo{shouldn't this be $\varphi^\dagger$?} Define $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ are the sequences on $\{(\vec z, \psi)\:|\:\psi\in \mathcal F, \vec z\in M^n\}$.
		\item We interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\vec z, (\vec z_1, \psi_1)\dots (\vec z_n, \psi_n)) = (\vec z_1, \psi_1)\dots (\vec z_n, \psi_n)(\vec z, \psi).$$
		\item For propositional variables $P$, we set $$P^{I_T}((\vec z_1, \psi_1)\dots (\vec z_n, \psi_n)) = P^I(f_{\psi_n}(\vec z_n, \dots(f_{\psi_1}(\vec z_1, s^I))\dots)).$$
	\end{itemize}
\end{definition}

\begin{lemma}
	If $\mathcal M$ is a counter-model to $\varphi^\circ$ \todo{shouldn't this be $\varphi^\dagger$?} then so is $\mathcal M_T$.
\end{lemma}

Even if we don't have the finiteness property of the propositional case, we have still managed to reduce the complexity of possible counter-models because we know that considering worlds of the form $$f_{\psi_1}(\vec z_1, f_{\psi_2}(\vec z_2, \dots f_{\psi_n}(\vec z_n, s)\dots))$$ is sufficient. 
In particular, this allows us to completely eliminate $\preceq$ from the formula, that is we remove the PartialOrder$(\preceq)$
from $K(\varphi)$ and replace $\forall u\forall w\forall x((E(x, u)\wedge u\preceq w)\to E(x, w))$ with $\bigwedge\{\forall \vec z\forall u(E(x, u)\to E(x, f_\psi(\vec z, u)))\:|\:\psi\in\mathcal F_\to\}$ and analogously for persistency to obtain $K$. We set

@@

\begin{lemma}\label{lemma:fo-simplification}
	$\varphi$ is intuitionistically valid if and only of $\varphi^\circ$ is classically valid.
\end{lemma}

\begin{proof}
	We proceed by translation of counter-examples. Suppose first we have a counter-example $\mathcal M = (M, I)$ to $\varphi^\circ$. As a Kripke frame $(W, \leq)$ take all $ W = \{m\in M\:|\:\text{ World}^I(m)\}$ let $\leq$ be $\preceq^I$ restricted to $W$. Then let $M_u = \{m\in M\:|\: E(m, u)\}$ and let $f^{I_u}$ be $f^I$ restricted to $M_u$ and $A^{I_u}(\vec x): \Leftrightarrow A^\#(\vec x, u)$. It is then a straightforward check of definitions that this defines a Kripke counter-model to $\varphi$.

	The other direction is a bit more involved. Suppose we have a Kripke counter-model to $\varphi$ with frame $(W, \preceq)$ and family of $\Sigma$-structures $(M_w, I_w)_{w\in W}$. In particular since it is a counter-model there exists $w_0\in W$ with $w_0\not\models\varphi$. Let $W_0 = \{w\in W\:|\: w\geq w_0\}$ and define an equivalence relation $\sim$ on $\{(x, u)\:|\:u\in W_0, x\in M_u\}$ via $(x, u)\sim (y, w)$ iff $x = y$ and there exists $v\in W_0$ comparable with both $u, w$ such that $x\in v$ and denote the equivalence class of $(x, u)$ with $[x, u]$. Let $M = W_0\cup \{[x, u]\:|\:u\in W_0, x\in M_u\}$.
	Now have
	\begin{itemize}
		\item $s^I = w_0$.
		\item $E^I(m, w)$ iff $w\in W_0$ and $m \sim [x, w]$ for some $x\in M_w$.
		\item $f^I(m_1\dots m_n) =\begin{cases}
			f^{I_u}(x_1\dots x_n), &\text{if there are $ u\in W_0, x_i\in M_u$ with $m_i\sim [x_i, u]$ for all $i$,}\\
			w_0, & \text{otherwise.}
		\end{cases}$
		\item ${A^\#}^I(m_1\dots m_n, u) \Leftrightarrow\begin{cases}
			A^{I_u}(x_1\dots x_n), &\text{if $u\in W_0$ and $\exists x_i\in M_u$ w. $m_i\sim [x_i, u]$ for all $i$,}\\
			\top, & \text{otherwise.}
		\end{cases}$
	\end{itemize}
	One easily verifies that these are well-defined. Let us now briefly check that this indeed defines a counter-model. First of all ${P^\#}^I(s^I)$ is false since $P^{I_{w_0}}$ is. Clearly World$(w_0)$ holds and $K(\varphi)$ is also easily verified. All that remains to show is $M, I\models\bigwedge\mathcal S^\circ$. Consider e.g. the case where $\psi\in\mathcal S$ is of the form
	$$ \forall \vec z((A(\vec a)\to B(\vec b))\to C(\vec c))$$
	We have to show that
	$$M, I\models \forall \vec z\forall u(\vec E(t, u)\to \forall w(u\preceq w\to A^\#(\vec a, w)\to B^\#(\vec b, w))\to C^\#(\vec c, u))$$
	holds. Suppose towards contradiction that it does not, i.e. there are $u, \vec z\in M$ such that for all $w\in M$
	$$\vec E^I(\vec z, u)\to (u\preceq w\to {A^\#}^I(\vec a, w)\to {B^\#}^I(\vec b, w))\to {C^\#}^I(\vec c, u)$$ is false. For that ${C^\#}^I(\vec c, u)$ must be false and in particular $u\in W_0$. Futhermore $\vec E^I(\vec z, u)$ must be true and we can write $\vec z = [z_1, u]\dots[z_n, u]$ for some $z_i\in W_u$. Note that the above is false in particular for $w\in M_0$ with $u\preceq w$. But this would imply
	$$((A^{I_u}(\vec a)\to B^{I_u}(\vec b))\to C^{I_u}(\vec c))[\vec z/z_1\dots z_n]$$in our original Kripke counter-model. The other cases are analogous.
\end{proof}

>>>>

\begin{definition}
Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^\circ$.
We say $\psi\in\mathcal F_\to$ is \emph{fulfilled} at $u\in M$ iff $A_\psi^I(u)\to B_\psi^I(u)$ is false or $C_\psi^I(u)$ is true. For $\psi\in\mathcal F_\to$ define $$g_\psi : M\to M, u\mapsto\begin{cases}
		u,&\text{ if $\psi$ is fulfilled in $u$,}\\
		f^I_\psi(u),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal F_\to$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $\psi_1\dots\psi_n$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}(\psi_1\dots \psi_n) = P^I(g_{\psi_n}(\dots(g_{\psi_1}(s^I))\dots)).$$
	\end{itemize}
\end{definition}
\begin{lemma}~\label{thm:prop-countermodel-reduction}
	Let $\mathcal M = (M, I)$ be a counter-model to $\mathcal \varphi^\circ$.
	\begin{enumerate}
		\item $\psi$ is fulfilled at $f_\psi^I(u)$ for all $u\in M, \psi\in\mathcal F_\to$.
		\item If $\psi$ is fulfilled at some $u\in M$ then $\psi$ is fulfilled at all $w\succeq u$.
	\end{enumerate}
\end{lemma}

@@

Using $E$ we then obtain $\mathcal S'$ analogously to the predicate case by including formulas:

@@

Then $\varphi = \bigwedge\mathcal S\to P$ is intuitionistically valid if and only if
$$\varphi' = \text{World}(s)\to K(\varphi)\to \bigwedge S'\to P^\#(s)$$
is classically valid, where $s$ is a new constant symbol. This was obvious in the propositional case but here it is more nuanced. A proof is given in the appendix as Lemma~\ref{lemma:fo-simplification}.

@@

As in the predicate case we now show the existence of a canonical counter-model (in case such a counter-model exists). Denote the set of formulas $\psi\in\mathcal S$ of the form $\forall\vec z((A(\vec a)\to B(\vec b))\to C(\vec c))$ with $\mathcal F_\to$ and of the form $\forall \vec z(\forall xA(\vec a)\to B(\vec b))$ with $\mathcal F_\forall$. Let $\mathcal F = \mathcal F_\to\cup\mathcal F_\forall$. For each  $\psi\in\mathcal F$ consider a new function symbol $f_\psi$. As before we may assume that $s$ is a least element and we can avoid the quantification over $w$ by considering the formulas
\begin{itemize}
	\item $\forall \vec z\forall u(\vec E(\vec z, u)\to (A^\#(\vec a, f_\psi(\vec z, u))\to B^\#(\vec b, f_\psi(\vec z, u)))\to C^\#(\vec c, u))$\\instead of $\forall \vec z\forall u(s\preceq u\to\vec E(\vec z, u)\to \forall w(u\preceq w\to A^\#(\vec a, w)\to B^\#(\vec b, w))\to C^\#(\vec c, u))$,
	\item $\forall \vec z\forall u(\vec E(\vec z, u)\to (\forall x(E(x, f_\psi(\vec z, u))\to A^\#(\vec a, f_\psi(\vec z, u))))\to B^\#(\vec b, u))$\\instead of $\forall \vec z\forall u(s\preceq u\to\vec E(\vec z, u)\to \forall w(u\preceq w\to \forall x(E(x, w)\to A^\#(\vec a, w)))\to B^\#(\vec b, u))$,
\end{itemize}
and by removing removing the $s\preceq u$ for the remaining formulas in $\mathcal S'$.
We denote the resulting set of formulas by $\mathcal S^\#$.
Then, the previous formulas are equivalid to
$$\varphi^\circ= \forall u(s\preceq u)\to \bigwedge\{\forall \vec t\forall u(u\preceq f(\vec z, u))\:|\:\psi\in\mathcal F\}\to K(\varphi)\to\bigwedge\mathcal S^\#\to P^\#(s).$$
As in the previous section we can define a tree counter-model for this.

\begin{definition}
	Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^\circ$. Define $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ are the sequences on $\{(\vec z, \psi)\:|\:\psi\in \mathcal F, \vec z\in M^n\}$.
		\item We interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\vec z, (\vec z_1, \psi_1)\dots (\vec z_n, \psi_n)) = (\vec z_1, \psi_1)\dots (\vec z_n, \psi_n)(\vec z, \psi).$$
		\item For propositional variables $P$, we set $$P^{I_T}((\vec z_1, \psi_1)\dots (\vec z_n, \psi_n)) = P^I(f_{\psi_n}(\vec z_n, \dots(f_{\psi_1}(\vec z_1, s^I))\dots)).$$
	\end{itemize}
\end{definition}

\begin{lemma}
	If $\mathcal M$ is a counter-model to $\varphi^\circ$ then so is $\mathcal M_T$.
\end{lemma}

Even if we don't have the finiteness property of the propositional case, we have still managed to reduce the complexity of possible counter-models because we know that considering worlds of the form $$f_{\psi_1}(\vec z_1, f_{\psi_2}(\vec z_2, \dots f_{\psi_n}(\vec z_n, s)\dots))$$ is sufficient. 
In particular, this allows us to completely eliminate $\preceq$ from the formula, that is we remove the PartialOrder$(\preceq)$
from $K(\varphi)$ and replace $\forall u\forall w\forall x((E(x, u)\wedge u\preceq w)\to E(x, w))$ with $\bigwedge\{\forall \vec z\forall u\forall x(E(x, u)\to E(x, f_\psi(\vec z, u)))\:|\:\psi\in\mathcal F_\to\}$ and analogously for persistency to obtain $K$. We set

@@

\begin{lemma}\label{lemma:fo-simplification}
	$\varphi$ is intuitionistically valid if and only of $\varphi'$ is classically valid.
\end{lemma}

\begin{proof}
	We proceed by translation of counter-examples. Suppose first we have a counter-example $\mathcal M = (M, I)$ to $\varphi'$. As a Kripke frame $(W, \leq)$ take all $ W = \{m\in M\:|\:\text{ World}^I(m)\}$ let $\leq$ be $\preceq^I$ restricted to $W$. Then let $M_u = \{m\in M\:|\: E(m, u)\}$ and let $f^{I_u}$ be $f^I$ restricted to $M_u$ and $A^{I_u}(\vec x): \Leftrightarrow A^\#(\vec x, u)$. It is then a straightforward check of definitions that this defines a Kripke counter-model to $\varphi$.

	The other direction is a bit more involved. Suppose we have a Kripke counter-model to $\varphi$ with frame $(W, \preceq)$ and family of $\Sigma$-structures $(M_w, I_w)_{w\in W}$. In particular since it is a counter-model there exists $w_0\in W$ with $w_0\not\models\varphi$. Let $W_0 = \{w\in W\:|\: w\geq w_0\}$ and define an equivalence relation $\sim$ on $\{(x, u)\:|\:u\in W_0, x\in M_u\}$ via $(x, u)\sim (y, w)$ iff $x = y$ and there exists $v\in W_0$ comparable with both $u, w$ such that $x\in v$ and denote the equivalence class of $(x, u)$ with $[x, u]$. Let $M = W_0\cup \{[x, u]\:|\:u\in W_0, x\in M_u\}$.
	Now have
	\begin{itemize}
		\item $s^I = w_0$.
		\item $E^I(m, w)$ iff $w\in W_0$ and $m \sim [x, w]$ for some $x\in M_w$.
		\item $f^I(m_1\dots m_n) =\begin{cases}
			f^{I_u}(x_1\dots x_n), &\text{if there are $ u\in W_0, x_i\in M_u$ with $m_i\sim [x_i, u]$ for all $i$,}\\
			w_0, & \text{otherwise.}
		\end{cases}$
		\item ${A^\#}^I(m_1\dots m_n, u) \Leftrightarrow\begin{cases}
			A^{I_u}(x_1\dots x_n), &\text{if $u\in W_0$ and $\exists x_i\in M_u$ w. $m_i\sim [x_i, u]$ for all $i$,}\\
			\top, & \text{otherwise.}
		\end{cases}$
	\end{itemize}
	One easily verifies that these are well-defined. Let us now briefly check that this indeed defines a counter-model. First of all ${P^\#}^I(s^I)$ is false since $P^{I_{w_0}}$ is. Clearly World$(w_0)$ holds and $K(\varphi)$ is also easily verified. All that remains to show is $M, I\models\bigwedge\mathcal S'$. Consider e.g. the case where $\psi\in\mathcal S$ is of the form
	$$ \forall \vec z((A(\vec a)\to B(\vec b))\to C(\vec c))$$
	We have to show that
	$$M, I\models \forall \vec z\forall u(\vec E(t, u)\to \forall w(u\preceq w\to A^\#(\vec a, w)\to B^\#(\vec b, w))\to C^\#(\vec c, u))$$
	holds. Suppose towards contradiction that it does not, i.e. there are $u, \vec z\in M$ such that for all $w\in M$
	$$\vec E^I(\vec z, u)\to (u\preceq w\to {A^\#}^I(\vec a, w)\to {B^\#}^I(\vec b, w))\to {C^\#}^I(\vec c, u)$$ is false. For that ${C^\#}^I(\vec c, u)$ must be false and in particular $u\in W_0$. Futhermore $\vec E^I(\vec z, u)$ must be true and we can write $\vec z = [z_1, u]\dots[z_n, u]$ for some $z_i\in W_u$. Note that the above is false in particular for $w\in M_0$ with $u\preceq w$. But this would imply
	$$((A^{I_u}(\vec a)\to B^{I_u}(\vec b))\to C^{I_u}(\vec c))[\vec z/z_1\dots z_n]$$in our original Kripke counter-model. The other cases are analogous.
\end{proof}
